Computing solution space properties by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-22 paper/paper.tex
Updates from Overleaf

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
This definition of generic programming covers two major aspects: ``work in the most general setting'' and ``with out loss of efficiency''.
By in the most general setting, we mean that a single program should work correctly for different input data types. For example, 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about the efficiency. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

@@

\section{Performance benchmarks}\label{sec:benchmark}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that have a small treewidth that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.
In this benchmark, we do not include traditional algorithms for finding the MIS size such as the branching algorithms~\cite{Tarjan1977, Robson1986} or dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013},
because to the best of our knowledge, most of the properties mentioned in this paper are not computable within these frameworks.
The main goal of this section is to show the relative timing for computing different properties.

@@

>>>>

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
This definition of generic programming covers two major aspects: ``work in the most general setting'' and ``with out loss of efficiency''.
By in the most general setting, we mean that a single program should work correctly for different input data types. For example, 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about the efficiency. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to graphics processing units (GPU) for faster computation~\cite{Besard2018}.

@@

\section{Performance benchmarks}\label{sec:benchmark}
We run a single thread benchmark on central processing units (CPU) Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that have a small treewidth that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.
In this benchmark, we do not include traditional algorithms for finding the MIS size such as the branching algorithms~\cite{Tarjan1977, Robson1986} or dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013},
because to the best of our knowledge, most of the properties mentioned in this paper are not computable within these frameworks.
The main goal of this section is to show the relative timing for computing different properties.

@@

To use it for solving solution space properties, just open a Julia REPL and type

\begin{lstlisting}
julia> using GraphTensorNetworks, Graphs

julia> solve(
           IndependentSet(
               Graphs.random_regular_graph(20, 3);
               optimizer = TreeSA(),
               weights = NoWeight(),
               openvertices = ()
           ),
           GraphPolynomial();
           usecuda=false
       )
0-dimensional Array{Polynomial{BigInt, :x}, 0}:
Polynomial(1 + 20*x + 160*x^2 + 659*x^3 + 1500*x^4 + 1883*x^5 + 1223*x^6 + 347*x^7 + 25*x^8)
\end{lstlisting}

Here the main function \texttt{solve} takes three inputs, the problem instance of type \texttt{IndependentSet}, the property instance of type \texttt{GraphPolynomial} and an optional key word argument \texttt{usecuda} to decide use GPU or not.
If one wants to use GPU to accelerate the computation, statement ``\texttt{using CUDA}'' must be executed before calling the solve function.
The problem instance takes four arguments to initialize, the only positional argument is the graph instance one wants to solve, the key word argument \texttt{optimizer} is for tensor network optimization, the key word argument \texttt{weights} specifies the weights of vertices as either a vector or \texttt{NoWeight()} and the keyword argument \texttt{openvertices} allows one to specify the degrees of freedom not summed over, i.e. returning a tensor rather than its sum.
Here, we use \texttt{TreeSA} method in Ref.~\cite{Kalachev2021}.
The first execution of the code will trigger Julia's just in time compiling,
which is a bit slow, but the following runs will be fast.
The following diagram lists possible combinations of input arguments, where graphs are mainly defined in package \texttt{Graphs}, and the rest can be found in \texttt{GraphTensorNetworks}.

\centerline{\includegraphics[width=0.8\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{figures/fig7.pdf}}
